---
title: "PS5-rmarkdown and plotting"
author: "Liu Yuliang"
date: "2/19/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
options(dplyr.summarize.inform = FALSE)
```
# Instructions
This is a problem set about rmarkdown and plotting (using ggplot). Unlike the previous problem sets, this one does not give you a ready-made GH repo with a code file–it is now your task to create a repo and include your rmarkdown file in there.\

You should answer the questions below in that file, knit it, and submit both the compiled html and link to your repo on canvas.\

- This problem sets asks you to write extensively when commenting your results. Please write clearly! Answer questions in a way that if the code chunks are hidden then the result is still readable! \
    * All substantial questions need explanations. You do not have to explain the simple things like “how many rows are there in data”, but if you make a plot of life expectancy, then you should explain what does the plot tell you. \
    * Write explanations as markdown and use the styles like bold and italic as appropriate.\
- Do not print too much results. It is all well to print a few lines of data for evaluation/demonstration purposes. But do not print dozens (or thousands!) of lines–no one bothers to look at that many numbers. You will lose points for annoying others (here your graders, but later potentially your boss). \
- Do not make code lines too long. 80-100 characters is a good choice. Your grader may not be able to follow all the code if the line is too long–most of us are using small laptop screens! (And again–you want to keep your graders happy!) \

# Gapminder data
We use gapminder dataset, downloaded from [**https://www.gapminder.org/data/**](https://www.gapminder.org/data/), however, the data structure there is quire complex, please use the dataset provided on canvas (in files/data). \
The variables are: \
**name** country name \
**iso3** 3-letter country code \
**iso2** 2-letter country code \
**region** broad geographic region \
**sub-region** more precise region \
**intermediate-region** \
**time** year \
**totalPopulation** total population \
**GDP_PC** GDP per capita (constant 2010 US$) \
**accessElectricity** Access to electricity (% of population) \
**agriculturalLand** Agricultural land (sq. km) \
**agricultureTractors** Agricultural machinery, tractors (count) \
**cerealProduction** Cereal production (metric tons) \
**feritilizerHa** Fertilizer consumption (kilograms per hectare of arable land) \
**fertilityRate** total fertility rate (births per woman) \
**lifeExpectancy** Life expectancy at birth, total (years) \
**childMortality** Mortality rate, under-5 (per 1,000 live births) \
**youthFemaleLiteracy** Literacy rate, youth female (% of females ages 15-24) \
**youthMaleLiteracy** Literacy rate, youth male (% of males ages 15-24) \
**adultLiteracy** Literacy rate, adult total (% of people ages 15 and above) \
**co2** CO2 emissions (kt) \
**greenhouseGases** Total greenhouse gas emissions (kt of CO2 equivalent) \
**co2_PC** CO2 emissions (metric tons per capita) \
**pm2.5_35** PM2.5 pollution, population exposed to levels exceeding WHO Interim Target-1 value 36ug/m3 \
**battleDeaths** Battle-related deaths (number of people) \

# 1 Load and check data (5pt)
You first task is to do a very simple data check: 

1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating). \
```{r}
gapminder <- read_delim("gapminder.csv")
```

2. (2pt) Load data. How many rows/columns do we have?  \
```{r}
num_of_rows <- nrow(gapminder)
num_of_rows
num_of_cols <- ncol(gapminder)
num_of_cols
```

3. (2pt) Print a small sample of data. Does it look OK? 
```{r}
gapminder %>% 
  head(5)
```
The data looks good, although there are many NA values.
  
  
---------------------------------------------------------------------


# 2 Descriptive statistics (15pt)

1. (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.
```{r}
iso3_count <- length(unique(gapminder$iso3))
iso2_count <- length(unique(gapminder$iso2))
name_count <- length(unique(gapminder$name))
iso3_count
iso2_count
name_count
```
There are `r iso3_count`, `r iso2_count` and `r name_count` countries in
iso3, iso2 and name, respectively.

2. If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}
multi_name_iso2 <- gapminder %>%
  filter(is.na(iso3) & is.na(name)) %>%
  group_by(iso2) %>%
  summarize(distinct_name = n_distinct(iso2)) %>% 
  print()
```
There is no iso-2 codes that correspond to more than one name.
    
    
    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.
```{r}
multi_name_name <- gapminder %>%
  filter(is.na(iso2) & is.na(iso3)) %>%
  group_by(name) %>%
  summarize(distinct_name = n_distinct(name)) %>% 
  print()

multi_name_iso3 <- gapminder %>%
  filter(is.na(iso2) & is.na(name)) %>%
  group_by(iso3) %>%
  summarize(distinct_name = n_distinct(iso3)) %>% 
  print()
```
There are 4 countries that have more than one iso3-code, which are CHANISL,
GBM, KOS and NLD_CURACAO.
    
3. (2pt) What is the minimum and maximum year in these data?
```{r}
min_year <- gapminder %>% 
  filter(!is.na(time)) %>% 
  summarize(min_year = min(time)) %>% 
  print()
max_year <- gapminder %>% 
  filter(!is.na(time)) %>% 
  summarize(max_year = max(time)) %>% 
  print()
```

---------------------------------------------------------------------

# 3 CO2 emissions (30pt)
Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data? \
```{r}
missing_co2 <- gapminder %>% 
  filter(!is.na(time)) %>% 
  group_by(time) %>% 
  summarize(missing_co2 = sum(is.na(co2))) %>%
  arrange(desc(missing_co2))

show_missing_co2 <- missing_co2 %>% 
  head() %>% 
  print()

missing_co2_PC <- gapminder %>% 
  filter(!is.na(time)) %>% 
  group_by(time) %>% 
  summarize(missing_co2_PC = sum(is.na(co2_PC))) %>%
  arrange(desc(missing_co2_PC))
show_missing_co2_PC <- missing_co2_PC %>% 
  head() %>% 
  print()
```
Year 2017, 2018 and 2019 have the most missing data that count for 217 NA values.

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see. \
```{r}
filter_countries_co2 <- gapminder %>% 
  filter(name == "United States of America" | name == "China" | name == "India"
         | name == "Germany" | name == "Iraq") %>% 
  filter(!is.na(co2))

ggplot(filter_countries_co2, aes(x = name, y = co2, fill = name)) +
  geom_col() +
  labs(title = "Total CO2 Emissions Over Time",
       x = "Year",
       y = "CO2 Emissions (kt)",
       fill = "Country")
```

United States of America has the most CO2 emissions, coming by China as the second place.
Iraq has the least CO2 emissions.

3. (5pt) Now let’s analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of thesame countries. What does this figure suggest? \
```{r}
filter_countries_co2_pc <- gapminder %>% 
  filter(name == "United States of America" | name == "China" | name == "India"
         | name == "Germany" | name == "Iraq") %>% 
  filter(!is.na(co2_PC))

ggplot(filter_countries_co2_pc, aes(x = name, y = co2_PC, fill = name)) +
  geom_col() +
  labs(title = "CO2 Emissions Per Capita Over Time",
       x = "Year",
       y = "CO2 Emissions (metric tons per capita)",
       fill = "Country")
```

Although United States of America still has the highest CO2 emissions per capita,
Germany becomes the second place and China drops to third. India becomes the lowest
country with CO2 emissions per capita among all five. The difference is due to 
the high population in China and India.

4. (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see. \
Note: just compute averages over countries and ignore the fact that countries are of different size. \
Hint: Americas 2016 should be 4.80. \
```{r}
overall_avg_co2 <- gapminder %>% 
  filter(!is.na(co2_PC) & !is.na(region)) %>%
  group_by(region) %>%
  summarize(overall_avg_co2 = mean(co2_PC, na.rm = TRUE)) %>% 
  print()

avg_co2_in_1960_and_2016 <- gapminder %>%
  filter(!is.na(co2_PC) & !is.na(region)) %>%
  filter(time == 1960 | time == 2016) %>%
  group_by(region, time) %>%
  summarize(avg_co2 = mean(co2_PC, na.rm = TRUE)) %>% 
  print()
```


5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016. \
Hint: it should look something along these lines(**Please refer to the spec for the graph**) \
```{r}
ggplot(avg_co2_in_1960_and_2016, aes(x = region, y = avg_co2, fill = factor(time))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Continent", y = "Average CO2 emissions per capita", fill = "Year") +
  ggtitle("Average CO2 Emissions per Capita Across Continents in 1960 and 2016")
```
The barplot shows the average CO2 emissions per capita across continents in 1960 and 2016.

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent). \
```{r}
top_co2_2019 <- gapminder %>% 
  filter(time == 2019) %>%
  filter(!is.na(region)) %>%
  group_by(region, name) %>% 
  summarize(mean_co2_PC = mean(co2_PC)) %>% 
  arrange(region, desc(mean_co2_PC)) %>% 
  top_n(3) %>% 
  print()

bot_co2_2019 <- gapminder %>% 
  filter(time == 2019) %>%
  filter(!is.na(region)) %>%
  group_by(region, name) %>% 
  summarize(mean_co2_PC = mean(co2_PC)) %>% 
  arrange(region, mean_co2_PC) %>% 
  top_n(3) %>% 
  print()
```

Since the data of CO2 per capita are missing in all 2019 year (all NA values),
the sorting of three largest and smallest CO2 emitters therefore yield no output.

---------------------------------------------------------------------

# 4 GDP per capita (50pt)
Let’s look at GDP per capita (*GDP_PC*)

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. \
Comment what do you see there. \
```{r}
data_1960 <- gapminder %>% 
  filter(time == 1960) %>% 
  filter(!is.na(GDP_PC) & !is.na(lifeExpectancy)) 

ggplot(data_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation)) +
    geom_point(aes(color = region)) + 
    labs(x = "GDP per Capita", y = "Life Expectancy by Country in Years",
         color = "Continents", size = "Total Population") +
    ggtitle("GDP per Capita versus Life Expectancy by Country in 1960")
```

Wealthier continents, like Americas, Europe and Oceania had a higher GDP per capita,
and a higher life expectancy. On the other hand, developing countries normally had
lower than 60 years of life expectancy.

2. (4pt) Make a similar plot, but this time use 2019 data only. \
```{r}
data_2019 <- gapminder %>% 
  filter(time == 2019) %>% 
  filter(!is.na(GDP_PC) & !is.na(lifeExpectancy)) 

ggplot(data_2019, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation)) +
    geom_point(aes(color = region)) + 
    labs(x = "GDP per Capita", y = "Life Expectancy by Country in Years",
         color = "Continents", size = "Total Population") +
    ggtitle("GDP per Capita versus Life Expectancy by Country in 2019")
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years? \

Wealthier continents, like Americas, Europe and Oceania still had a higher GDP per capita,
and a higher life expectancy after 60 years. However, developing countries had
an roughly increase of 10 years of life expectancy, adding to over 60 years in many African countries, and
about 70 years for most Asian countries. This shows that the developing countries had a steady improvement on
medical services, food and water quality, etc., so that people can live longer.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? \
Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.\
```{r}
avg_life_exp_in_1960_and_2016 <- gapminder %>%
  filter(!is.na(lifeExpectancy) & !is.na(region)) %>%
  filter(time == 1960 | time == 2016) %>%
  group_by(region, time) %>%
  summarize(avg_life_exp = mean(lifeExpectancy, na.rm = TRUE)) %>% 
  print()
```
The result fits in the relationship in the figure, as developed countries in Oceania, Europe and Americas still
have higher life expectancy, but developing countries in Africa and Asia are catching up quickly.

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see. \
Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\
```{r}
avg_le_growth <- gapminder %>% 
  filter(!is.na(lifeExpectancy) & !is.na(region)) %>%
  filter(time <= 2019 & time >= 1960) %>%
  group_by(region) %>%
  mutate(prev = lag(lifeExpectancy),
         growth = (lifeExpectancy - prev)) %>%
  summarize(avg_le_growth = mean(growth, na.rm = TRUE)) %>%
  arrange(avg_le_growth) %>%
  print()
```
The average life expectancy growth is high in developing countries in Asia and Africa,
and has a lower growth in developed countries in Oceania, Europe and Americas.

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it! \
```{r}
GDP_PC_1960_to_2019 <- gapminder %>% 
  filter(time == 1960 | time == 2019) %>% 
  filter(!is.na(GDP_PC))

ggplot(GDP_PC_1960_to_2019, aes(y = GDP_PC)) +
  geom_histogram(aes(fill = factor(time))) +
  labs(title = "GDP per capita in 1960 and 2019", y = "GDP per capita", fill = "Year")
```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.) \
Hint: check out the function rank()! \

```{r}
rank_in_1960 <- gapminder %>% 
  filter(time == 1960) %>% 
  arrange(desc(lifeExpectancy)) %>% 
  mutate(rank_in_1960 = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  select(rank_in_1960) %>% 
  print()

rank_in_1960 <- gapminder %>% 
  filter(time == 2019) %>% 
  arrange(desc(lifeExpectancy)) %>% 
  mutate(rank_in_2019 = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America") %>% 
  select(rank_in_2019) %>% 
  print()
```


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? \
Hint: 0.0904 for 1960. \
```{r}
LE_in_1906_and_2019 <- gapminder %>% 
  filter(!is.na(lifeExpectancy) & !is.na(region)) %>%
  filter(time == 1960 | time == 2019)
num_countries_1960 <- LE_in_1906_and_2019 %>% 
  filter(time == 1960) %>% 
  select(lifeExpectancy) %>% 
  nrow()
num_countries_2019 <- LE_in_1906_and_2019 %>% 
  filter(time == 2019) %>% 
  select(lifeExpectancy) %>% 
  nrow()
relative_rank_1960 <- 18 / num_countries_1960
relative_rank_2019 <- 47 / num_countries_2019
relative_rank_1960
relative_rank_2019
```
The relative rank in 1960 was 0.0957, and 0.2398 in 2019.

---------------------------------------------------------------------

Finally tell us how many hours did you spend on this PS.

6 hours
